---
title: "CD45+Thy1+ ILC transcription factor activity analysis"
output:
   html_document: 
     theme: united
     df_print: kable
     toc: yes
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include = FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```


# Universal packages

```{r Load packages}
library(Seurat)
library(tidyverse)
library(patchwork)
library(cowplot)
```


# Data input

```{r Load SCT processed data}
# Load ILC data

# ILC
ILC <- readRDS("CD45_Thy1_ILC_paper_ver.rds")


# Set RNA assay as default
DefaultAssay(ILC) <- "RNA"
head(x = ILC[[]])
```


# ILC cluster Visualization

```{r Set Reveal ILC clusters}
# Show the numbers for ILC
table(ILC@meta.data$cell_type, split.by = ILC@meta.data$treatment)

# Plot the ILC clusters
DimPlot(ILC, label = T, repel =  T)
DimPlot(ILC, group.by = "treatment", label = F) + ggtitle(NULL)
```


# Packages for transcription factor activity inference  analysis

```{r Load package for transcription factor activity analysis}
# Load the package (decoupleR)
library(decoupleR)

# Only needed for data handling and plotting
library(pheatmap)
```


# TF activity inference with Weighted Mean

```{r Load CollecTRI network}
# Load the database of TFs and their transcriptional targets
net <- get_collectri(organism='mouse', split_complexes=FALSE)

head(net)
```

```{r Calculate the pseudotime on averall ILC}
# Run the Weighted Mean method (wmean)
# Extract the normalized log-transformed counts
ILC.mat <- as.matrix(ILC@assays$RNA@data)

# Run wmean
ILC.acts <- run_wmean(mat=ILC.mat, net=net, .source='source', .target='target',
                  .mor='mor', times = 100, minsize = 5)
head(ILC.acts)

# Extract norm_wmean and store it in tfswmean
ILC[['tfswmean']] <- ILC.acts %>%
  filter(statistic == 'norm_wmean') %>%
  pivot_wider(id_cols = 'source', names_from = 'condition',
              values_from = 'score') %>%
  column_to_rownames('source') %>%
  Seurat::CreateAssayObject(.)

# Change assay
DefaultAssay(object = ILC) <- "tfswmean"

# Scale the data (ILC)
ILC <- ScaleData(ILC)
ILC@assays$tfswmean@data <- ILC@assays$tfswmean@scale.data
```


# Split of the data by treatment

```{r Split the data}
ILC <- SetIdent(ILC, value = "treatment")

# BPC
BPC <- ILC %>% 
  subset(idents = c("Alt.BPC"), invert = FALSE) %>%
  SetIdent(value = "cell_type")

table(BPC@meta.data$cell_type, split.by = BPC@meta.data$treatment)


# Alt
Alt <- ILC %>% 
  subset(idents = c("Alt"), invert = FALSE) %>%
  SetIdent(value = "cell_type")

table(Alt@meta.data$cell_type, split.by = Alt@meta.data$treatment)
```


# Exploration of the overall ILC TF activities

```{r TF analysis shared settings}
# Select top variable TFs
n_tfs <- 60

# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-3, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 3, length.out=floor(palette_length/2)))
```

```{r Overall ILC TF exploration split by treatments, fig.width = 12, fig.height = 3}
# Extract activities from object as a long dataframe
ILC.df.treatment <- t(as.matrix(ILC@assays$tfswmean@data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(ILC)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Get top tfs with more variable means across clusters
ILC.tfs.treatments <- ILC.df.treatment %>%
  group_by(source) %>%
  summarise(std = sd(mean)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)

# Subset long data frame to top TFs and transform to wide matrix (by treatments)
top_ILC.acts_mat_treatment <- ILC.df.treatment %>%
  filter(source %in% ILC.tfs.treatments) %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()


# Plot by treatment
pheatmap(top_ILC.acts_mat_treatment, border_color = NA, color=my_color, breaks = my_breaks)
```

```{r Overall ILC TF exploration split by cell types, fig.width = 12, fig.height = 4}
# Set the primary ident to "cell_type"
ILC <- SetIdent(ILC, value = "cell_type")

# Extract activities from object as a long dataframe
ILC.df.cell_type <- t(as.matrix(ILC@assays$tfswmean@data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(ILC)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Get top tfs with more variable means across clusters
ILC.tfs.cell_type <- ILC.df.cell_type %>%
  group_by(source) %>%
  summarise(std = sd(mean)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)

# Subset long data frame to top TFs and transform to wide matrix (by cell types)
top_ILC.acts_mat_cell_type <- ILC.df.cell_type %>%
  filter(source %in% ILC.tfs.cell_type) %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()


# Plot by cell types
pheatmap(top_ILC.acts_mat_cell_type, border_color = NA, color=my_color, breaks = my_breaks)
```


# Exploration of the Alt TF activities

```{r Alt TF exploration, fig.width = 12, fig.height = 4}
# Extract activities from object as a long dataframe
Alt.df <- t(as.matrix(Alt@assays$tfswmean@data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(Alt)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Get top tfs with more variable means across clusters
Alt.tfs <- Alt.df %>%
  group_by(source) %>%
  summarise(std = sd(mean)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)

# Subset long data frame to top TFs and transform to wide matrix
top_Alt.acts_mat <- Alt.df %>%
  filter(source %in% Alt.tfs) %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()


# Plot
pheatmap(top_Alt.acts_mat, border_color = NA, color=my_color, breaks = my_breaks) 
```

```{r Alt Ahr expression , fig.width = 13, fig.height = 4}
Alt_Ahr_p1 <- DimPlot(Alt, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
Alt_Ahr_p2 <- (FeaturePlot(Alt, features = c("Ahr")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('AhR activity in Alt')
DefaultAssay(object = Alt) <- "RNA"
Alt_Ahr_p3 <- FeaturePlot(Alt, features = c("Ahr")) + ggtitle('Ahr expression in Alt')
DefaultAssay(object = Alt) <- "tfswmean"

plot_grid(Alt_Ahr_p1, Alt_Ahr_p2, Alt_Ahr_p3, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```

```{r Alt Irf1 expression, fig.width = 13, fig.height = 4}
Alt_Irf1_p1 <- DimPlot(Alt, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
Alt_Irf1_p2 <- (FeaturePlot(Alt, features = c("Irf1")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('IRF1 activity in Alt')
DefaultAssay(object = Alt) <- "RNA"
Alt_Irf1_p3 <- FeaturePlot(Alt, features = c("Irf1")) + ggtitle('Irf1 expression in Alt')
DefaultAssay(object = Alt) <- "tfswmean"

plot_grid(Alt_Irf1_p1, Alt_Irf1_p2, Alt_Irf1_p3, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```

```{r Alt Myc expression, fig.width = 13, fig.height = 4}
Alt_Myc_p1 <- DimPlot(Alt, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
Alt_Myc_p2 <- (FeaturePlot(Alt, features = c("Myc")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('MYC activity in Alt')
DefaultAssay(object = Alt) <- "RNA"
Alt_Myc_p3 <- FeaturePlot(Alt, features = c("Myc")) + ggtitle('Myc expression in Alt')
DefaultAssay(object = Alt) <- "tfswmean"

plot_grid(Alt_Myc_p1, Alt_Myc_p2, Alt_Myc_p2, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```


# Exploration of the Alt.BPC TF activities

```{r BPC TF exploration, fig.width = 12, fig.height = 4}
# Extract activities from object as a long dataframe
BPC.df <- t(as.matrix(BPC@assays$tfswmean@data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(BPC)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Get top tfs with more variable means across clusters
BPC.tfs <- BPC.df %>%
  group_by(source) %>%
  summarise(std = sd(mean)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)

# Subset long data frame to top TFs and transform to wide matrix
top_BPC.acts_mat <- BPC.df %>%
  filter(source %in% BPC.tfs) %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()


# Plot
pheatmap(top_BPC.acts_mat, border_color = NA, color=my_color, breaks = my_breaks)
```

```{r BPC Ahr expression , fig.width = 13, fig.height = 4}
BPC_Ahr_p1 <- DimPlot(BPC, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
BPC_Ahr_p2 <- (FeaturePlot(BPC, features = c("Ahr")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('AhR activity in Alt.BPC')
DefaultAssay(object = BPC) <- "RNA"
BPC_Ahr_p3 <- FeaturePlot(BPC, features = c("Ahr")) + ggtitle('Ahr expression in Alt.BPC')
DefaultAssay(object = BPC) <- "tfswmean"

plot_grid(BPC_Ahr_p1, BPC_Ahr_p2, BPC_Ahr_p3, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```

```{r BPC Irf1 expression , fig.width = 13, fig.height = 4}
BPC_Irf1_p1 <- DimPlot(BPC, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
BPC_Irf1_p2 <- (FeaturePlot(BPC, features = c("Irf1")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('IRF1 activity in Alt.BPC')
DefaultAssay(object = BPC) <- "RNA"
BPC_Irf1_p3 <- FeaturePlot(BPC, features = c("Irf1")) + ggtitle('Irf1 expression in Alt.BPC')
DefaultAssay(object = BPC) <- "tfswmean"

plot_grid(BPC_Irf1_p1, BPC_Irf1_p2, BPC_Irf1_p3, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```

```{r BPC Myc expression , fig.width = 13, fig.height = 4}
BPC_Myc_p1 <- DimPlot(BPC, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
BPC_Myc_p2 <- (FeaturePlot(BPC, features = c("Myc")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('MYC activity in Alt.BPC')
DefaultAssay(object = BPC) <- "RNA"
BPC_Myc_p3 <- FeaturePlot(BPC, features = c("Myc")) + ggtitle('Myc expression in Alt.BPC')
DefaultAssay(object = BPC) <- "tfswmean"

plot_grid(BPC_Myc_p1, BPC_Myc_p2, BPC_Myc_p3, nrow = 1, rel_widths = c(1, 1.3, 1.3))
```


# Packages for external picture input

```{r Load Image-Processing package}
# Load Image-Processing package (magick)
library(magick)
```

```{r Save the pheatmap}
# Save the pheatmap
pheatmap(top_BPC.acts_mat, border_color = NA, color=my_color, breaks = my_breaks, cellwidth = 14, cellheight = 14, fontsize = 10, main = 'Transcription Factor activity in Alt.BPC', filename = "BPC_TFs.pdf")
```

```{r Convert pdf to png}
# Load pdf tp png conversion package 
library(pdftools)

pdf_convert(
  "./BPC_TFs.pdf",
  format = "png",
  pages = NULL,
  filenames = "BPC_TFs.png",
  dpi = 600,
  antialias = TRUE,
  opw = "",
  upw = "",
  verbose = TRUE
)
```


# Data oupput

```{r TF activity for paper, fig.width = 12, fig.height = 10}
# Input the conclusion picture
BPC_TFs <- image_read("./BPC_TFs.png")

figure7_A <- ggdraw() +
  draw_image(BPC_TFs, scale = 1)


# Generate the plot in paper figure form
figure7_1 <- plot_grid(figure7_A, labels = c('(A)'), label_size = 18)
figure7_2 <- plot_grid(Alt_Irf1_p2, Alt_Myc_p2, Alt_Ahr_p2, BPC_Irf1_p2, BPC_Myc_p2, BPC_Ahr_p2, labels = c('(B)', '(D)', '(F)', '(C)', '(E)', '(G)'), label_size = 18, nrow = 2, ncol = 3, rel_widths = c(1, 1, 1))

figure7 <- plot_grid(figure7_1, figure7_2,
                     nrow = 2,
                     ncol = 1,
                     rel_heights = c(1, 2)
  )

figure7
```

```{r Conclusion figure, fig.width = 6, fig.height = 6}
conclusion <- image_read("./conclusion.png")

figureS4 <- ggdraw() +
  draw_image(conclusion, scale = 1)

figureS4
```


# Data save

```{r Output the figures}
# Create a folder for the figures
dir.create("./Figures", showWarnings = FALSE)

# Save the figure
ggsave2(plot = figure7, path = "./Figures", "Figure7.tiff", device = "tiff", dpi = 300, bg = "white", width = 11, height = 9, units = "in")
ggsave2(plot = figureS4, path = "./Figures", "FigureS4.tiff", device = "tiff", dpi = 300, bg = "white", width = 6, height = 6, units = "in")
```

```{r Output the CSV files}
# Create a folder for CSV files
dir.create("./CSV", showWarnings = FALSE)

# Export to csv
write.csv(ILC.df.treatment, file.path("./CSV", "Alt+Alt.BPC_by_treatment_ILC_TF_activities.csv"))
write.csv(ILC.df.cell_type, file.path("./CSV", "Alt+Alt.BPC_by_cell_type_ILC_TF_activities.csv"))
write.csv(BPC.df, file.path("./CSV", "Alt.BPC_ILC_TF_activities.csv"))
write.csv(Alt.df, file.path("./CSV", "Alt_ILC_TF_activities.csv"))
```


<details>
  <summary>**Session Info**</summary>
```{r sessionInfo}
sessionInfo()
```
</details>

